Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-10 paper_io_sim.tex
replaced tabs with spaces
Added Chris as co-author (nighres application)

\author{
  \IEEEauthorblockN{
    Hoang-Dung Do\IEEEauthorrefmark{1},
    Val\'erie Hayot-Sasson\IEEEauthorrefmark{1},
    Rafael Ferreira da Silva\IEEEauthorrefmark{3},
    Henri Casanova\IEEEauthorrefmark{2},
    Tristan Glatard\IEEEauthorrefmark{1}
  }\\
  \IEEEauthorblockA{
    \IEEEauthorrefmark{1}Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
    \IEEEauthorrefmark{2}Department of Information and Computer Sciences, University of Hawai`i at M\=anoa, USA\\
    \IEEEauthorrefmark{3}Information Sciences Institute, University of Southern California, Marina Del Rey, CA, USA
  }
}

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, writethrough enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends writethrough
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform writethrough as well as
            direct I/O as it delays disk writes to perform a bulk write at a later time
            \cite{linuxdev3rd2010}.

            Cache eviction and flushing strategies are integral to proper page cache functioning.
            Whenever space in memory becomes limited, either as a result of application memory
            or page cache use, page cache data may be evicted. Only data that
            has been persisted to storage (clean pages) can be flagged for eviction and removed from
            memory. Written data that has not yet been persisted to disk (dirty data) must first
            be copied (flushed) to storage prior to eviction. When sufficient memory is
            being occupied, the flushing process is synchronous. However, even when
            there is sufficient available memory, written data will be flushed to disk
            at a predefined interval through a process known as \textit{periodical flushing}.
            Periodical flushing only flushes expired dirty pages, which remain dirty in
            page cache longer than an expiration time configured in the kernel.
            Different cache eviction algorithms have also been developed and proposed
            \cite{owda2014comparison}.
            The Linux kernel uses a two-list strategy to flag pages for eviction.
            The two-list strategy is based on a least recently used (LRU) policy
            and uses an active and inactive list in its implementation.
            If accessed pages are not in the page cache, they are added to the inactive list.
            Should pages located on the inactive list be accessed, they will be moved from
            the inactive to the active list.
            The lists are also kept balanced by moving pages from the active list
            to the inactive list when the active list grows too large.
            Thus, the active list only contains pages which are accessed more than once
            and not evictable, while the inactive list includes pages accessed once only,
            or pages that have be accessed more than once but moved from the active list.
            Both lists operate using LRU eviction policies meaning the data that has
            not be accessed recently will be moved first.

>>>>

\author{
  \IEEEauthorblockN{
    Hoang-Dung Do\IEEEauthorrefmark{1},
    Val\'erie Hayot-Sasson\IEEEauthorrefmark{1},
    Rafael Ferreira da Silva\IEEEauthorrefmark{3},
    Christopher Steele\IEEEauthorrefmark{4}
    Henri Casanova\IEEEauthorrefmark{2},
    Tristan Glatard\IEEEauthorrefmark{1}
  }\\
  \IEEEauthorblockA{
    \IEEEauthorrefmark{1}Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
    \IEEEauthorrefmark{2}Department of Information and Computer Sciences, University of Hawai`i at M\=anoa, USA\\
    \IEEEauthorrefmark{3}Information Sciences Institute, University of Southern California, Marina Del Rey, CA, USA\\
    \IEEEauthorrefmark{4}Department of Psychology, Concordia University, Montreal, Canada
  }
}

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
        The goal of the page cache is to try ensure that most I/O operations occur entirely
        within memory space rather than synchronously to and from the storage device.
        All disk-backed data located in the page cache are partitioned into many small pages.
        When a read operation is initiated, the kernel checks whether the required pages are in memory.
        If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
        the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
        Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
        writethrough and writeback.
        If direct~I/O is enabled for a given file system, all I/O operations bypass the
        page cache for writes, writing data directly to disk. In contrast, writethrough enables the page cache
        for file system writes, but the data is written synchronously from memory to disk. Writeback extends writethrough
        by enabling asynchronous writes to disk, meaning that a write call will return once
        the data has been written to memory.
        The writeback strategy is considered to outperform writethrough as well as
        direct I/O as it delays disk writes to perform a bulk write at a later time
        \cite{linuxdev3rd2010}.
        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes expired dirty pages, which remain dirty in
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been developed and proposed
        \cite{owda2014comparison}.
        The Linux kernel uses a two-list strategy to flag pages for eviction.
        The two-list strategy is based on a least recently used (LRU) policy
        and uses an active and inactive list in its implementation.
        If accessed pages are not in the page cache, they are added to the inactive list.
        Should pages located on the inactive list be accessed, they will be moved from
        the inactive to the active list.
        The lists are also kept balanced by moving pages from the active list
        to the inactive list when the active list grows too large.
        Thus, the active list only contains pages which are accessed more than once
        and not evictable, while the inactive list includes pages accessed once only,
        or pages that have be accessed more than once but moved from the active list.
        Both lists operate using LRU eviction policies meaning the data that has
        not be accessed recently will be moved first.
